---
title: "Google Capstone Project"
subtitle: "Cyclistic bike-share analysis case study! "
author: "Emmanuel Ncube"
date: "`r Sys.Date()`"
output: html_document
---

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

## INTRODUCTION

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

#### This project is part of requirements in fulfillment of the Google Data Analysis Professional Certificate.

#### The purpose of this script is to consolidate downloaded Divvy data for the year 2021 into a single data frame and then conduct an analysis to help answer the key question: "In what ways do members and casual riders use Divvy bikes differently?"

### Setting up the environment

```{r install the packages}

# install.packages("tidyverse") # for data wrangling
# install.packages("janitor")   # for data cleaning
# install.packages("lubridate") # for date manipulation
# install.packages("vroom")     
# install.packages("ggplot2")   # for data visualization
# 
# getwd() # prints the current working directory
# setwd("C:/Users/hp/Desktop/Google Capstone") # set the working directly to simplify data reference
```

### Load packages

```{r libraries}
library(tidyverse)
library(janitor)
library(lubridate)
```

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

# STEP 1: COLLECT DATA

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

### Lets read the data and make columns consistent

```{r cleaning column names }

# create a list of .csv files
file_list <- list.files(path = "C:/Users/hp/Desktop/Google Capstone/divvy_raw", 
             pattern="*.csv", full.names = TRUE)
df <- lapply(file_list, read.csv) %>% bind_rows()
glimpse(df)
```

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

# STEP 2: DATA WRANGLING & TRANSFORMATION

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

### Remove start_lat, start_lng,end_lat and end_lng columns as these columns are irrelevant to the current analysis

```{}
```

```{r remove irrelevant columns}
all_trips <- select(df, -c(start_lat,start_lng,end_lat,end_lng))
```

### Inspect the new table created

```{r inspect columns}
colnames(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
```

### There are a few issues with the data that needs fixing:

### 1. Some columns of data need to be added e.g.. day, month, year, to allow aggregation of the data.

### 2. Adding of length of ride column to the entire data frame to allow bike usage analysis.

### Let's add columns that list the date, month, day and year of each ride

```{r year,month,day,day of the week}

all_trips$year <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$year), "%m")
all_trips$day <- format(as.Date(all_trips$year), "%d")
all_trips$day_of_week <- format(as.Date(all_trips$year), "%A")
head(all_trips)
```

# Add a "ride_length" calculated field (in seconds)

```{r add ride_length column}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

# Lets inspect the structure of the columns

```{r file inspection}
str(all_trips)
```

### Let's remove "Bad data" and create a new version of the data frame (v2) e.g. rides with negative ride length

```{r remove negative ride length}

all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]
```

### Check the data structure again

```{r inspect the file}
# str(all_trips_v2)
```

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

# STEP 3: CONDUCT DESCRIPTIVE ANALYSIS

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

## Summary statistics by membership

### Let's order the weekdays to enable grouping by weekday

```{r ordering weekdays}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

### Summary statistics (max, mean of ride length and mode of the weekday)

```{r summary statistics}
all_trips_v2 %>%
  group_by(member_casual) %>% 
  summarise(max_ride_length = max(ride_length),
            mean_ride_length = mean(ride_length),
            mode_weekday = which.max(table(day_of_week)))
```

### Average ride_length by membership and weekday

```{r avg ride duration by membership and weekday}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% # group by member_casual and weekday
  summarise(average_ride_length = mean(ride_length)) # no. of rides and average duration
```

### Let's see ridership by membership type and weekday

```{r avg ride duration by membership and weekday}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% # group by member_casual and weekday
  summarise(number_of_rides = n()) %>% 
  mutate("proportion_membership (%)" = round((number_of_rides/sum(number_of_rides)*100)))
```

### Let's plot the number of rides per day for each membership

```{r}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% # group by member_casual and weekday
  summarise(number_of_rides = n()) %>% # no. of rides and average duration
  ggplot(aes(x = day_of_week ,y = number_of_rides, fill=member_casual)) +
  geom_col(position="dodge")
```

### Let's create a visualization for average ride_length by weekday and membership

```{r}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% # group by member_casual and weekday
  summarise(number_of_rides = n(), 
            avg_ride_length = mean(ride_length)) %>% # no. of rides and average duration
  ggplot(aes(x=day_of_week,y = avg_ride_length, fill=member_casual)) + geom_col(position="dodge")
```

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

# STEP 4: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS

### \*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

### Create a csv file that we will visualize in Excel, Tableau or my presentation software

### Exporting of file for further analysis

```{r data to be exported}
all_trips_v2 <- all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% # group by member_casual and weekday
  summarise(average_ride_length = mean(ride_length), total_rides = n())

type(all_trips_v2) # check type of object (its a list)
```

### Let's convert the list to data frame for easy exporting to a csv file:

```{r converting list to data frame}

# Lets convert to a data frame and export the data to csv file
trips_final <- data.frame(all_trips_v2)
write_csv(trips_final, file = 'C:\\Users\\hp\\Desktop\\Google Capstone\\Divvy_Exercise\\avg_ride_length.csv')
```

### Let's check how the csv will look like:

```{r}
f <- read.csv("avg_ride_length.csv", sep = ";")
View(f)
```

# KEY FINDINGS

1.  Riders with memberships are more than 50% of the total riders.
2.  Casual riders tend to hire more of bikes from Friday right through the weekend.
3.  This suggests that casual riders use bikes for leisure, tourism or exercising.
4.  Members continue to use bikes to go to work observed in constant ride length throughout the week (fig. 1).

# RECOMMENDATIONS

1.  Introducing weekend membership at a discount

2.  Introduce discounts for longer riders only if they are members

3.  Charging casual riders per ride length and discounts for longer riders i.e. Ride More Pay Less!
